Exploring the Demographic Aspects of COVID-19 Cases in Toronto: Are the elderly males more susceptible to COVID-19?

### 2025-06-17 outputs/paper/paper.qmd
changed scripts

```{r}
#| include: false
#| warning: false
#| message: false

# install.packages("tidyverse")
# install.packages("here")
# install.packages("knitr")

library(tidyverse)
library(here)
library(knitr)

# Set the working directory to the project's root
setwd(here())

### Explore ###
# Read cleaned case data
cleaned_case_data <- 
  read_csv(
    "outputs/data/cleaned_case_data.csv", 
    show_col_types = FALSE)

# Read analyzed case data
analyzed_case_data <-
  read_csv(
    "outputs/data/analyzed_case_data.csv",
    show_col_types = FALSE)
```

@@

```{r}
#| label: tbl-idcasetable
#| tbl-cap: Sample of cleaned COVID-19 cases data
#| echo: false
#| message: false
#| warning: false

# Display the first six rows of the "cleaned_case_data" dataset
head(cleaned_case_data) |>
  kable(
    # Give each column a name
    col.names = c("Case ID", "Age group", "Gender")
  )
```

@@

```{r}
#| label: fig-casebygender
#| fig-cap: Number of covid-19 cases by gender
#| echo: false
#| warning: false
#| message: false

# Create a bar chart to visualize the distribution of COVID-19 cases by gender based on the analyzed_case_data dataset
# Reference: https://tellingstorieswithdata.com/02-drinking_from_a_fire_hose.html#explore 
analyzed_case_data |> 
  ggplot(aes(x = gender)) +
  # Change bar color to green
  geom_bar(fill = "#D4E7C5") +
  geom_text(
    # Use the 'count' statistic to calculate counts for each category
    stat = 'count',
    # Use percentage labels based on count and set y position of text labels
    aes(label = scales::percent(..count.. / sum(..count..)),
        y = ..count..),
    # Adjust vertical position of text labels
    vjust = -0.5,
    # Stack text labels on top of bars
    position = position_stack(vjust = 0.5)
  ) +
  # Apply a minimal theme to the chart
  theme_minimal() +
  # Label x-axis as "Gender" and y-axis as "Number of COVID-19 cases"
  labs(x = "Gender",
       y = "Number of COVID-19 cases")
```

@@

```{r}
#| label: fig-casebyagegroup
#| fig-cap: Number of covid-19 cases by age group
#| echo: false
#| warning: false
#| message: false

# Create a bar chart to visualize the distribution of COVID-19 cases by age group based on the analyzed_case_data dataset
# Reference: https://tellingstorieswithdata.com/02-drinking_from_a_fire_hose.html#explore 
analyzed_case_data |> 
  ggplot(aes(x = age_group)) +
  # Change bar color to green
  geom_bar(fill = "#D4E7C5") +
  geom_text(
    # Use the 'count' statistic to calculate counts for each category
    stat = 'count',
    # Use percentage labels based on count and set y position of text labels
    aes(label = scales::percent(..count.. / sum(..count..)),
        y = ..count..),
    # Adjust vertical position of text labels
    vjust = -0.5,
    # Stack text labels on top of bars
    position = position_stack(vjust = 0.5),
    # Change text size to 3
    size = 3
  ) +
  # Apply a minimal theme to the chart
  theme_minimal() +
  # Label x-axis as "Age group" and y-axis as "Number of COVID-19 cases"
  labs(x = "Age group",
       y = "Number of COVID-19 cases")
```

@@

```{r}
#| label: fig-distribution
#| fig-cap: Distribution of covid-19 cases across genders and age groups
#| echo: false
#| warning: false
#| message: false

# Reference: https://rstudio-pubs-static.s3.amazonaws.com/215516_c2cee6b2885d419e9248e7eb7c87f887.html
analyzed_case_data |> 
  # Create a ggplot object with age_group on the x-axis and gender as fill color
  ggplot(aes(x = age_group, fill = gender)) +
  # Add a grouped bar layer to create the grouped bar chart
  geom_bar(position = "dodge") +
  # Scale the x-axis labels to display "Age group"
  scale_x_discrete("Age group") +
  # Scale the fill color to display "Gender" and specify the break points: "Male", "Female", "Other"
  scale_fill_discrete(name = "Gender",
                      breaks = c("Male", "Female", "Other")) +
  # Apply a minimal theme to the chart
  theme_minimal() +
  # Label y-axis as "Number of COVID-19 cases"
  labs(y = "Number of COVID-19 cases")
```

>>>>

```{python}
#| include: false
#| warning: false
#| message: false

#### Workspace setup ####
import pandas as pd

#### Explore ####
# Read cleaned case data
cleaned_case_data = pd.read_csv("../data/cleaned_case_data.csv")
analyzed_case_data = pd.read_csv("../data/analyzed_case_data.csv")

```

@@

```{python}
#| label: tbl-idcasetable
#| tbl-cap: Sample of cleaned COVID-19 cases data
#| echo: false
#| message: false
#| warning: false

sample = cleaned_case_data.head().rename(
    columns={
        "case_id": "Case ID",
        "age_group": "Age group",
        "gender": "Gender"
    }
)

sample
```

@@

```{python}
#| label: fig-casebygender
#| fig-cap: Number of COVID-19 cases by gender
#| echo: false
#| warning: false
#| message: false

import matplotlib.pyplot as plt

# Count occurrences by gender
gender_counts = analyzed_case_data["gender"].value_counts()
gender_percent = gender_counts / gender_counts.sum()

# Plot
fig, ax = plt.subplots()

bars = ax.bar(gender_counts.index, gender_counts.values, color="#D4E7C5")

# Add percentage labels above bars
for i, (count, pct) in enumerate(zip(gender_counts.values, gender_percent)):
    ax.text(i, count + 1000, f"{pct:.1%}", ha='center', va='bottom')

# Label axes and apply minimal formatting
ax.set_xlabel("Gender")
ax.set_ylabel("Number of COVID-19 cases")
ax.set_title("")  # Leave empty if you don't want a title
ax.spines['top'].set_visible(False)
ax.spines['right'].set_visible(False)
ax.yaxis.grid(True, linestyle='--', alpha=0.4)
ax.set_axisbelow(True)

plt.tight_layout()
plt.show()

```

@@

```{python}
#| label: fig-casebyagegroup
#| fig-cap: Number of COVID-19 cases by age group
#| echo: false
#| warning: false
#| message: false

import matplotlib.pyplot as plt

# Count occurrences by age group and preserve order
age_order = ["0-19", "20-29", "30-39", "40-49", "50-59", "60-69", "70-79", "80-89", "90+"]
age_counts = analyzed_case_data["age_group"].value_counts().reindex(age_order)
age_percent = age_counts / age_counts.sum()

# Plot
fig, ax = plt.subplots(figsize=(10, 5))

bars = ax.bar(age_counts.index, age_counts.values, color="#D4E7C5")

# Add percentage labels
for i, (count, pct) in enumerate(zip(age_counts.values, age_percent)):
    ax.text(i, count + max(age_counts.values) * 0.01, f"{pct:.1%}", ha='center', va='bottom', fontsize=8)

# Label axes and apply minimal styling
ax.set_xlabel("Age group")
ax.set_ylabel("Number of COVID-19 cases")
ax.spines['top'].set_visible(False)
ax.spines['right'].set_visible(False)
ax.yaxis.grid(True, linestyle='--', alpha=0.4)
ax.set_axisbelow(True)

plt.tight_layout()
plt.show()

```

@@

```{python}
#| label: fig-distribution
#| fig-cap: Distribution of COVID-19 cases across genders and age groups
#| echo: false
#| warning: false
#| message: false

import matplotlib.pyplot as plt
import numpy as np

# --- Prep counts -----------------------------------------------------------
age_order    = ["0-19", "20-29", "30-39", "40-49", "50-59",
                "60-69", "70-79", "80-89", "90+"]
gender_order = ["Male", "Female", "Other"]

counts = (
    analyzed_case_data
      .groupby(["age_group", "gender"])
      .size()                               # tally
      .unstack(fill_value=0)                # pivot to wide
      .reindex(age_order)                   # keep age order
)[gender_order]                              # keep gender order

# --- Plot ------------------------------------------------------------------
x      = np.arange(len(age_order))
width  = 0.25                               # width of each bar
fig, ax = plt.subplots(figsize=(10, 5))

for i, gender in enumerate(gender_order):
    ax.bar(x + (i - 1)*width,               # dodge position
           counts[gender],
           width=width,
           label=gender)

# Axis / style tweaks
ax.set_xlabel("Age group")
ax.set_ylabel("Number of COVID-19 cases")
ax.set_xticks(x)
ax.set_xticklabels(age_order)
ax.legend(title="Gender")
ax.spines["top"].set_visible(False)
ax.spines["right"].set_visible(False)
ax.yaxis.grid(True, linestyle="--", alpha=0.4)
ax.set_axisbelow(True)

plt.tight_layout()
plt.show()

```
